---
title: "Group Project Stats 112"
author: "Alec, Jay, Lynn"
date: "5/7/2021"
output: html_document
---

##Setup ---

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(nlme)
library(mgcv)
aids = read.csv("aids.csv")

glimpse(aids)
```

##Exploratory Data Analysis --- 

```{r}
subjects = unique(aids$id)
subjects = max(subjects)
```

The number of subjects in the dataset is `r subjects`
The number of the covariates is `r ncol(aids) `

##Univariate summaries 
```{r}
# Distribution of Age
aids %>% 
  ggplot(aes(x = age)) + 
  geom_histogram(bins = 15, color = "black", fill = "gray")
```


```{r}
# Distribution of people in each treatment group
table(aids$treatment)
round(prop.table(table(aids$treatment)), 2)
```



##Bivariate summaries 
```{r}
#Boxplot based on treatment group
aids %>% 
  ggplot(aes(x = factor(treatment), y = log_cd4, 
             fill = factor(treatment))) +
  geom_boxplot()
```

```{r}
#Boxplot based on gender
aids %>% 
  ggplot(aes(x = gender, y = log_cd4, fill = gender)) +
  geom_boxplot()
```


##Overall trends of response ---

```{r}
# Mean log_cd4 per week of each treatment 

aids %>% 
 group_by(treatment, week) %>% 
 summarize(mean_logcd4 = mean(log_cd4)) %>% 
 ggplot(aes(x = week, y = mean_logcd4,  color =  factor(treatment))) +  
 labs(title = "Smoothed Trend of Each Treatment",
      y = "Mean Log cd4", x = "Week") +
 geom_smooth(se = FALSE) +
  facet_wrap(~treatment)
```

```{r}

# Trends pf each subject per week separated by treatment
aids %>% 
 ggplot(aes(x = week, y = log_cd4, color = factor(treatment)))+
 geom_point(alpha = 0.3) +
 geom_smooth(se = FALSE) +
 labs(title = "Smoothed Trend of Each Subject",
      y = " Log cd4", x = "Week")
```

```{r}

#Trends by grouping by treatment and week
aids %>% 
  group_by(treatment, week) %>% 
  summarize(mean_log = mean(log_cd4)) %>% 
  ggplot(aes(x = week, y = mean_log, color = factor(treatment))) +
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE)
```


##Imbalances in the data ---





## Outliers ---

We can see from the boxplots above there are quite a few outliers in the log_cd4 per treatment. To see the overall number of outliers we can create a boxplot of the log_cd4 and count the numbers outside of the interquartile range


```{r}
aids %>% 
  ggplot(aes(y = log_cd4)) +
  geom_boxplot()
```
These are the outliers and the total number of outliers are 
```{r}
#outliers values
tibble(outliers = boxplot.stats(aids$log_cd4)$out )

#number of outliers 
observerd = tibble(r = boxplot.stats(aids$log_cd4)$out) %>% 
  nrow()

#expected number of outliers are
expected = nrow(aids) * .05
```
There is about `r observerd` outliers and the expected is `r expected`.
We are below the number of expected number of outliers so the observed outliers are normal.


<p>Let us check the distribution of the log_cd4 and assess its normality</p>
```{r}

aids %>% 
  ggplot(aes(x = log_cd4)) +
   geom_histogram(bins = 15, color = "black", fill = "gray") +
   labs(title = "Histogram of Log_cd4")
  

```

The histogram its approximately normal, but we can get more information but standardizing the log_cd4

```{r}

# Standardized of log_cd4 with mean 0 and sd 1 
log_cd4 = aids$log_cd4

scaled = scale(log_cd4)
colMeans(scaled)
apply(scaled, 2, sd)

tibble(r = scaled) %>% 
  ggplot(aes(x = r)) +
  geom_histogram( color = "black", fill = "gray") +
  labs(title = "Standardized Histogram of Log_cd4")
```

We can see that there is quite a few number of instances after the -2. This indicates heavy tails as well as outliers
